ANNO: The Anthropological Notation Ontology

### 2023-03-31 paper.tex
Add module and SMOG images.

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
\usepackage{todonotes}
%\usepackage{dcolumn}

@@

% KH: Ich hab hier mal den Text vom Poster als Basis für die Introduction genommen
\section{Introduction}\label{sec:introduction}
Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
Osteometry, describes quantitative measurement on the human skeletton.
Anthropology relies on osteometric measurements which has several problems:
Bones can only be investigated in place, which requires transport over far distances.
This causes wear in the bones, and can lead to different measurements.
This can be alleviated with a digital twin, which can be measured at any location.

@@

The \emph{Terminologia anatomica}~\citep{ta2} (TA) is a hierarchy of anatomical landmarks of the entire human body.
From its history, the Latin and Greek designations have emerged to allow standardization.
It is advantageous that these languages no longer show progress and are increasingly losing political power. 
In TA, the order of structures follows the natural anatomy.
An anatomical landmark is identified by an individual identification number, the anatomical term, and the English term.
The identification number is critical because certain landmarks are redundant.
For example, there is the \emph{processus zygomaticus} at the \emph{os frontale} and the \emph{os temporale}.
Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
Terms included in the \emph{other} category given include eponyms, which are concepts named after people, places or things.
For example, \emph{ossa digitorum pedis} (engl. \emph{phalanges of the foot}) are the small bones of the toes and are also denoted as \emph{ossa phalangea pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
% chapter 2: Ossa (Bones)

The \emph{Foundational Model of Anatomy} (FMA) ontology~\citep{fma} represents the physical organization of human anatomy.
%It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
However, the Latin and Greek designations are not taken into account, due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
This has a negative impact on the comprehensibility for users.

@@

% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
% TODO: ontology language, expressivity
% TODO: reuse of ontological entities
% TODO: clear validation mechanism
% TODO: top-level ontology?
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, which is transformed to an OWL 2 ontology.
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.

@@

While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.
ANNO links to concepts of the FMA when they exist, but structures them in a different way.
The ontology consists of the following core modules:
%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
\subsection{Bones}

@@

\section{Use Case: Integration into \aw{}}
% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
ANNO is used by \aw{}~\citep{anthroworks3d}.
By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
This enables location-independent and parallel work without wear and tear on the bone.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.

>>>>

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
\usepackage{todonotes}
\usepackage{cleveref}
%\usepackage{dcolumn}

@@

% KH: Ich hab hier mal den Text vom Poster als Basis für die Introduction genommen
\section{Introduction}\label{sec:introduction}
Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
Osteometry describes quantitative measurement on the human skeletton.
Anthropology relies on osteometric measurements, which has several problems:
Bones can only be investigated in place, which requires transport over far distances.
This causes wear in the bones, and can lead to different measurements.
This can be alleviated with a digital twin, which can be measured at any location.

@@

The \emph{Terminologia anatomica}~\citep{ta2} (TA) is a hierarchy of anatomical landmarks of the entire human body.
From its history, the Latin and Greek designations have emerged to allow standardization.
It is advantageous that these languages no longer show progress and are increasingly losing political power. 
In TA, the order of structures follows the natural anatomy.
An anatomical landmark is identified by an individual identification number, the anatomical term, and the English term.
The identification number is critical because certain landmarks are redundant.
For example, there is the \emph{processus zygomaticus} at the \emph{os frontale} and the \emph{os temporale}.
Synonyms, like \emph{zygomatic bone} and \emph{malar bone}, are also given.
Terms included in the \emph{other} category include eponyms, which are concepts named after people, places or things.
For example, \emph{ossa digitorum pedis} (engl. \emph{phalanges of the foot}) are the small bones of the toes and are also denoted as \emph{ossa phalangea pedis} which are named after the Greek word \emph{phalanx} for a dense rectangular infantry formation.
% chapter 2: Ossa (Bones)

The \emph{Foundational Model of Anatomy} (FMA) ontology~\citep{fma} represents the physical organization of human anatomy.
%It allows the knowledge it contains to be represented in a human-understandable and machine-interpretable way.
However, it does not include Greek and Latin labels, which are frequently used by medical professionals.
Due to hypernyms 90\,\% of the terms are re-designated and only 1\,\% contain definitions.
This has a negative impact on the comprehensibility for users.

@@

% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
% TODO: ontology language, expressivity
% TODO: reuse of ontological entities
% TODO: clear validation mechanism
% TODO: top-level ontology?
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}
The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.\todo{explain}

@@

While there are many textual sources of anthropological systematization, they do not agree in all aspects and there is not a single, formally described, standard.
ANNO links to concepts of the FMA when they exist, but structures them in a different way.

\begin{figure}[h]
\includegraphics[width=0.3\textwidth]{img/modules.pdf}
\caption{The six modules of ANNO.}\label{fig:modules}
\end{figure}

The ontology consists of the following core modules, see \cref{fig:modules}:
%For the categories, there were the following spreadsheets: bones, anatomical landmarks, osteometric landmarks, measured distances, function (separately for discriminant and regress functions), anatomical position descriptions, sources used, and osteological terminology.
\subsection{Bones}

@@

\section{Use Case: Integration into \aw{}}
% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
ANNO is used by \aw{}~\citep{anthroworks3d}.
By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
This enables location-independent and parallel work without wear and tear on the bone.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:

\begin{itemize}
\item ontology-oriented generation of placeholders/containers for objects to be imported
\item variable properties for different bone elements and bone types 
\item hierarchy for orientation within an examination project according to the ontology tree
\item categorization of measurements and markers according to specifications from the ontology
\item generation of input forms based on the specified properties and bone types
\item automatic generation of measurements
\end{itemize}
